SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-12-04 design_arch.tex
Updated 3.1 to be centered around figure discussion

As previously discussed, ultimately it might be profitable to allow a given
prefetcher to speculate as deeply as possible, since often some useful
prefetches are generated long after the confidence of the prefetcher has
fallen below the point at which performance improvements are seen due to the
increasing rate of inaccurate prefetches.  However, to allow this deep
speculation in the prefetcher the increasing numbers of bad prefetches must be
filtered out.  Thus, we propose to leverage perceptron-based learning as a
mechanism to differentiate between potentially useful deeply speculated
prefetches and likely not-useful ones.  This Perceptron Prefetch Filter (PPF)
is placed between the prefetcher and the prefetch insertion queue, to ensure
that predicted bad prefetches are not inserted into the memory system.
\pg{As shown in figure whatever...}.

@@

% Gino: Are these modifications minimal?
% EB: That's a good question :P. Edited to 'appropriate'.

@@

%Gino: rephrase below?
The Perceptron Filter is organized as tables of perceptron weights, with a
separate table of weights for each feature. For a design using `N' number of
features, its weights will be arranged in $N$ different tables. Each feature's
value is used to index into the corresponding table.  Since each table is
indexed independently, different numbers of bits can be used for indexing,
thus giving each table a different size.  The variable indexing can be
determined by studying the features and fine-tuning them empirically to
achieve a good accuracy versus hardware overhead trade-off.  Exact details of
our PPF implementation are given under ``Area Overheads'' in
Section~\ref{Method-Overheads}.

%\pg{this whole ``bits'' discussion belongs in implementation not here IMHO}
%[EB] Resolved

A single entry in the table corresponds to a perceptron weight.  Each weight
is a 5 bit counter saturating at -16 and +15 and initialized to zero at the
beginning of a program's execution. We observed empirically that 5 bits of
index achieves a good trade-off between hardware overheads and accuracy.

%\pg{the fact that its nine is again implementation details}
% [EB] Resolved

% \pg{The word ChampSim should not appear in this part of the paper at
%   all.  Remember you are describing how to build this in the general
%   case...} 
% [EB] Resolved
\textbf{Inference}\newline The prefetcher is triggered on every L2 cache
demand access.  At that point, the base prefetcher has the option to perform a
prefetch -- if it does, then it has a choice of how many cache lines to
prefetch.  The suggested prefetches can be either placed in the L2 cache or
the last-level cache based on the prefetcher's internal confidence mechanism.

When the base prefetcher is triggered, it begins to suggest candidates for
prefetching.  All suggested prefetches are passed through the perceptron
filter to decide whether they are qualified for final prefetching.  The filter
decides whether to prefetch a candidate by looking at the microarchitectural
state, \textit{i.e.} the features, at that instant.  Each feature is hashed to
form an index into a table of up to 4,096 entries dedicated to that feature.

Once all the weights are retrieved, they are summed and compared to a preset
threshold, PERC\_ THRESHOLD\_LO.  Only prefetch candidates with perceptron sum
higher than the threshold qualify for prefetching.  The prefetches that
qualify through the perceptron stage are recorded in the ``Prefetch Table''.
The prefetch table is a 1,024-entry, direct mapped structure that contains all
metadata required to re-index the perceptron entries for training.  When the
feedback of the current prefetched line is available at a later stage, the
stored data is used to train the perceptron.

%\pg{this discussion below has redundancy and doesn't flow well.
%  Rework it}

Each feature's confidence contributes to the final decision of whether to
prefetch a cache line, with the summation of all features' confidences
denoting the overall confidence for a prefetch.  The final sum is compared
against two thresholds: PERC\_THRESHOLD\_HI and  PERC\_ THRESHOLD\_ LO to
decide the placement of the data in the cache hierarchy. Prefetches that have
a confidence above the higher threshold are placed into the L2 cache since the
observed features strongly recommend  the prefetch. A prefetch that results in
a confidence between PERC\_ THRESHOLD\_ LO and PERC\_ THRESHOLD\_ HI are
placed in the larger last level cache, as the filter is moderately confident
of the cache line's use, but seeks to reduce pollution in the smaller L2
cache. Finally, the remaining prefetches that are below the lower threshold
are not prefetched due to a lack of correlation between the current observed
features and the prediction from the base prefetch engine.

%GAC: The section's previous contents are below, in case the above 
%rewritten work is not what we were looking for in terms of a rewording
%
%The perceptron sum can be considered the sum of the individual
%contribution of a feature.  The value of each contribution corresponds
%to a feature's confidence in the final decision on whether to prefetch
%a cache line. By summing the individual contributions, the final
%perceptron sum denotes the overall confidence for that prefetch
%suggestion.  By comparing the perceptron sum to two different
%thresholds, we divide the confidence scale of the prefetch suggestion
%into three bins. The first bin corresponds to the lowest confidence
%and leads to prefetch candidate being rejected.  The next bin
%corresponds to prefetches with a moderate confidence level.  This bin
%prefetches directly to fill the bigger last level cache and prevents
%polluting the smaller L2 Cache.  The highest confidence prefetches are
%filled in the L2 cache.
%
%
%The sum of the perceptron weights are also used to decide whether to
%place the data in the L2 or the last level cache.  All prefetch
%candidates that are qualified until this stage are compared against
%PERC\_THRESHOLD\_HI to decide the fill level. The two thresholds:
%PERC\_THRESHOLD\_\_LO and PERC \_THRESHOLD\_HI are empirically set.

In addition to the prefetch table mentioned above, PPF also maintains a
``Reject Table.''  The reject table is a 1,024-entry direct-mapped table.
If a prefetch suggestion is rejected by the perceptron layer, it is logged
into the reject table.  The table is used to train the perceptron to avoid
false negatives \textit{i.e.}, cases where prediction was to reject the
prefetch but the prefetch would have been useful.

%As previously stated all the metadata describing the program state at the
%instant of prefetching is stored in the Prefetch Table or the Reject Table.
%This information is useful when the perceptron needs to be updated
%subsequently.

\textbf{Training}\newline In the prefetching environment, feedback for a
prefetch is received whenever there is an eviction or a demand access from the
L2 cache.  This action triggers training of the base prefetcher as well as the
perceptron layer.  Training the perceptrons involves accessing the metadata
that was stored in Prefetch Table or the Reject Table.  The cache lines
address of the block being trained is used to access a table, using 10 bits of
the address for indexing and another 6 bits to perform tag matching.  Once the
values of the features at the time of prefetching is available, it is used to
index into the perceptron weights table.

If the demanded block that triggers the training was tagged as a valid
prefetch in the prefetch table, then the earlier prefetch prediction was
correct.  In that case the perceptron weights are incremented by 1 if the
predicted sum does not cross a pre-defined threshold. These training
thresholds are introduced to avoid overfitting of the perceptron weights to
the given program behavior. These thresholds are referred to as
POS\_UPDATE\_THRESHOLD ($\theta_p$) and NEG\_UPDATE\_THRESHOLD ($\theta_n$),
respectively for the positive and negative values of training saturation.

@@

>>>>

\begin{figure}
  \begin{center}
  \includegraphics[width=9cm]{PPF_Hierarchy.pdf}
    \label{fig:PPF_Hierarchy}
  \caption{PPF Architecture in the Memory Hierarchy}
  \end{center}
\end{figure}

As previously discussed, ultimately it might be profitable to allow a given
prefetcher to speculate as deeply as possible, since often some useful
prefetches are generated long after the confidence of the prefetcher has
fallen below the point at which performance improvements are seen due to the
increasing rate of inaccurate prefetches.  However, to allow this deep
speculation in the prefetcher the increasing numbers of bad prefetches must be
filtered out.  Thus, we propose to leverage perceptron-based learning as a
mechanism to differentiate between potentially useful deeply speculated
prefetches and likely not-useful ones.  This Perceptron Prefetch Filter (PPF)
is placed between the prefetcher and the prefetch insertion queue, to ensure
that predicted bad prefetches are not inserted into the memory system as shown in 
figure~\ref{fig:PPF_Hierarchy}.

@@

@@

The Perceptron Filter and the steps used to performed inference 
and training are shown in figure~\ref{fig:PPF_Datapath}. The filter is organized as tables of 
perceptron weights, with a separate table of weights for each feature. 
For a design using `N' number of features, its weights will be arranged
in $N$ different tables. Each feature's value is used to index 
into the corresponding table with $K$ entries.  Since each table is
indexed independently, different numbers of bits can be used for indexing,
thus giving each table a different size.  The variable indexing can be
determined by studying the features and fine-tuning them empirically to
achieve a good accuracy versus hardware overhead trade-off.  Exact details of
our PPF implementation are given under ``Area Overheads'' in
Section~\ref{sec:Method-Overheads}.

A single entry in the table corresponds to a perceptron weight.  Each weight
is a 5 bit counter saturating at -16 and +15 and initialized to zero at the
beginning of a program's execution. We empirically observe that 5 bits of
index achieves a good trade-off between hardware overhead and accuracy.\newline

\begin{figure}
  \begin{center}
  \includegraphics[width=9cm]{Datapath.pdf}
    \label{fig:PPF_Datapath}
  \caption{PPF Data Path and Operation}
  \end{center}
\end{figure}

\noindent{\textbf{1. Inference}}\newline

The prefetcher is triggered on every L2 cache
demand access.  At that point, the base prefetcher has the option to perform a
prefetch -- if it does, then it has a choice of how many cache lines to
prefetch.  The suggested prefetches can be either placed in the L2 cache or
the last-level cache based on the prefetcher's internal confidence mechanism.

When the base prefetcher is triggered, it begins to suggest candidates for
prefetching.  All suggested prefetches pass through the perceptron
filter to decide whether they are qualified for final prefetching.  The filter
decides whether to prefetch a candidate by looking at the microarchitectural
state, \textit{i.e.} the features, at that instant.  Each feature is hashed to
form an index into a table of up to 4,096 entries dedicated to that feature.

Each feature's confidence contributes to 
the final decision of whether to prefetch a cache line, with the summation 
of all features' confidences denoting the overall confidence for a prefetch.  
The final sum is compared against two thresholds: PERC\_THRESHOLD\_HI and  
PERC\_ THRESHOLD\_ LO to decide the placement of the data in the cache 
hierarchy. Prefetches that have a confidence above the higher threshold 
are placed into the L2 cache since the observed features strongly recommend  
the prefetch. A prefetch that results in a confidence between 
PERC\_ THRESHOLD\_ LO and PERC\_ THRESHOLD\_ HI are placed in the larger 
last level cache, as the filter is moderately confident of the cache line's 
use, but seeks to reduce pollution in the smaller L2 cache. Finally, the 
remaining prefetches that are below the lower threshold are not prefetched 
due to a lack of correlation between the current observed features and 
the prediction from the base prefetch engine.\newline

\noindent{\textbf{2. Recording}}\newline

The prefetches that qualify through the inference stage are recorded in the 
``Prefetch Table''. The prefetch table is a 1,024-entry, direct mapped 
structure that contains all metadata required to re-index the perceptron 
entries for training.

In addition to the prefetch table mentioned above, 
PPF also maintains a 1,024-entry direct-mapped ``Reject Table.'' 
If a prefetch suggestion is rejected by the perceptron layer, it is logged into 
the reject table. The table is used to train the perceptron to avoid false negatives 
\textit{i.e.}, cases where prediction was to reject the prefetch but 
the prefetch would have been useful based on the observed demand accesses to the L2. \newline

\noindent{\textbf{3. Feedback and Data Retrieval}} \newline

In the prefetching environment, feedback for a prefetch is received whenever 
there is an eviction or a demand access from the L2 cache. This action 
triggers training of the base prefetcher as well as the perceptron layer.  
The cache line's address of the block being trained is used to access either the Prefetch
or Reject table, using 10 bits of the address for indexing and another 6 bits to 
perform tag matching.  Once the values of the features at the time of prefetching is available,
it is used to index into the perceptron weights table. \newline

\noindent{\textbf{4. Training}}\newline

Training the perceptrons involves accessing the metadata
that was stored in Prefetch Table or the Reject Table.

If the demanded block that triggers the training was tagged as a valid
prefetch in the prefetch table, then the earlier prefetch prediction was
correct.  In that case the perceptron weights are incremented by 1 if the
predicted sum does not cross a pre-defined threshold. These training
thresholds are introduced to avoid overfitting of the perceptron weights to
the given program behavior. These thresholds are referred to as
POS\_UPDATE\_THRESHOLD ($\theta_p$) and \newline
NEG\_UPDATE\_THRESHOLD ($\theta_n$),
respectively for the positive and negative values of training saturation.

@@

%Gino: rephrase below?
%The Perceptron Filter is organized as tables of perceptron weights, with a
%separate table of weights for each feature. For a design using `N' number of
%features, its weights will be arranged in $N$ different tables. Each feature's
%value is used to index into the corresponding table.  Since each table is
%indexed independently, different numbers of bits can be used for indexing,
%thus giving each table a different size.  The variable indexing can be
%determined by studying the features and fine-tuning them empirically to
%achieve a good accuracy versus hardware overhead trade-off.  Exact details of
%our PPF implementation are given under ``Area Overheads'' in
%Section~\ref{Method-Overheads}.
%
%%\pg{this whole ``bits'' discussion belongs in implementation not here IMHO}
%%[EB] Resolved
%
%A single entry in the table corresponds to a perceptron weight.  Each weight
%is a 5 bit counter saturating at -16 and +15 and initialized to zero at the
%beginning of a program's execution. We observed empirically that 5 bits of
%index achieves a good trade-off between hardware overheads and accuracy.
%
%%\pg{the fact that its nine is again implementation details}
%% [EB] Resolved
%
%% \pg{The word ChampSim should not appear in this part of the paper at
%%   all.  Remember you are describing how to build this in the general
%%   case...} 
%% [EB] Resolved
%
%\begin{figure}
%  \begin{center}
%  \includegraphics[width=9cm]{Datapath.pdf}
%    \label{fig:PPF_Datapath}
%  \caption{PPF Data Path and Operation}
%  \end{center}
%\end{figure}
%
%\textbf{Inference}\newline The prefetcher is triggered on every L2 cache
%demand access.  At that point, the base prefetcher has the option to perform a
%prefetch -- if it does, then it has a choice of how many cache lines to
%prefetch.  The suggested prefetches can be either placed in the L2 cache or
%the last-level cache based on the prefetcher's internal confidence mechanism.
%
%When the base prefetcher is triggered, it begins to suggest candidates for
%prefetching.  All suggested prefetches are passed through the perceptron
%filter to decide whether they are qualified for final prefetching.  The filter
%decides whether to prefetch a candidate by looking at the microarchitectural
%state, \textit{i.e.} the features, at that instant.  Each feature is hashed to
%form an index into a table of up to 4,096 entries dedicated to that feature.
%
%Once all the weights are retrieved, they are summed and compared to a preset
%threshold, PERC\_ THRESHOLD\_LO.  Only prefetch candidates with perceptron sum
%higher than the threshold qualify for prefetching.  The prefetches that
%qualify through the perceptron stage are recorded in the ``Prefetch Table''.
%The prefetch table is a 1,024-entry, direct mapped structure that contains all
%metadata required to re-index the perceptron entries for training.  When the
%feedback of the current prefetched line is available at a later stage, the
%stored data is used to train the perceptron.
%
%%\pg{this discussion below has redundancy and doesn't flow well.
%%  Rework it}
%
%Each feature's confidence contributes to the final decision of whether to
%prefetch a cache line, with the summation of all features' confidences
%denoting the overall confidence for a prefetch.  The final sum is compared
%against two thresholds: PERC\_THRESHOLD\_HI and  PERC\_ THRESHOLD\_ LO to
%decide the placement of the data in the cache hierarchy. Prefetches that have
%a confidence above the higher threshold are placed into the L2 cache since the
%observed features strongly recommend  the prefetch. A prefetch that results in
%a confidence between PERC\_ THRESHOLD\_ LO and PERC\_ THRESHOLD\_ HI are
%placed in the larger last level cache, as the filter is moderately confident
%of the cache line's use, but seeks to reduce pollution in the smaller L2
%cache. Finally, the remaining prefetches that are below the lower threshold
%are not prefetched due to a lack of correlation between the current observed
%features and the prediction from the base prefetch engine.
%
%In addition to the prefetch table mentioned above, PPF also maintains a
%``Reject Table.''  The reject table is a 1,024-entry direct-mapped table.
%If a prefetch suggestion is rejected by the perceptron layer, it is logged
%into the reject table.  The table is used to train the perceptron to avoid
%false negatives \textit{i.e.}, cases where prediction was to reject the
%prefetch but the prefetch would have been useful.
%
%%As previously stated all the metadata describing the program state at the
%%instant of prefetching is stored in the Prefetch Table or the Reject Table.
%%This information is useful when the perceptron needs to be updated
%%subsequently.
%
%\textbf{Training}\newline In the prefetching environment, feedback for a
%prefetch is received whenever there is an eviction or a demand access from the
%L2 cache.  This action triggers training of the base prefetcher as well as the
%perceptron layer.  Training the perceptrons involves accessing the metadata
%that was stored in Prefetch Table or the Reject Table.  The cache lines
%address of the block being trained is used to access a table, using 10 bits of
%the address for indexing and another 6 bits to perform tag matching.  Once the
%values of the features at the time of prefetching is available, it is used to
%index into the perceptron weights table.
%
%If the demanded block that triggers the training was tagged as a valid
%prefetch in the prefetch table, then the earlier prefetch prediction was
%correct.  In that case the perceptron weights are incremented by 1 if the
%predicted sum does not cross a pre-defined threshold. These training
%thresholds are introduced to avoid overfitting of the perceptron weights to
%the given program behavior. These thresholds are referred to as
%POS\_UPDATE\_THRESHOLD ($\theta_p$) and NEG\_UPDATE\_THRESHOLD ($\theta_n$),
%respectively for the positive and negative values of training saturation.
%
%If a cache block eviction led to training and the corresponding valid entry is
%found in the prefetch table without the block being accessed, then the
%prediction made by the perceptron was wrong.  The perceptron should have
%ideally rejected the prefetch suggestion as a low-confidence prefetch.  Here
%the weights are decremented by 1 to reflect the misprediction. In either case,
%weights are saturated at -16 or +15.
%
%A secondary training mechanism also kicks in during demand fetches.  Before
%the demand access triggers the next set of prefetches, the reject table is
%checked for a valid entry.  A hit means that the corresponding cache line was
%initially suggested by the underlying prefetch engine, but rejected by the
%perceptron filter.  Thus, the perceptron should have been more confident about
%that particular prefetch.  Once such a scenario is identified, the values of
%the features at the time of prefetch is retrieved from the reject table.  The
%retrieved data is used to index into the various weights tables of the
%perceptron and the corresponding values are updated by +1, saturating between
%and -16 and +15.  This update reflects increased confidence for the prediction
%corresponding to that prefetch.
%
%This mechanism allows us to exploit a previously lost opportunity.  In prior
%perceptron based implementations, and in general, prefetching algorithms,
%there is usually no way of knowing the result of not prefetching a particular
%line.  Our two-step PPF architecture allows us to overcome that issue.
